ProteInfer: deep networks for protein functional inference

### 2021-10-06 latex/draft3.tex
Updates

Beyond functional annotation, deep learning has enabled significant advances in protein structure prediction \cite{alquraishi2019end, senior2020improved, Yang1496, yang2020improved, du2020energy, rao2021msa}, predicting the functional effects of mutations \cite{riesselman2018deep,rives2019biological,rao2019evaluating,frazer2020large}, and protein design \cite{yang2019machine, mazurenko2019machine, biswas2020low, madani2020progen,elnaggar2020prottrans, anishchenko2020novo, bryant2021deep}. A key departure from traditional approaches is that researchers have started to incorporate vast amounts of raw, uncurated sequence data into model training.

@@

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are alternative methods for splitting, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. Such a split is not available for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives \citep{warwick2020benchmarking}.

@@

%startleaveout
\begin{acknowledgements}
  We would like to thank Babak Alipanahi, Jamie Smith, Eli Bixby, Drew Bryant, Shanqing Cai, Cory McLean and Abhinay Ramparasad. The static version of this manuscript uses a template made by Ricardo Henriques.
  \end{acknowledgements}
  \begin{contributions}
  TS was responsible for conception, data ingestion, code, analysis, visualisation, UI, paper writing. MLB was
  responsible for visualization, analysis, code, CLI, open sourcing, testing, paper writing. DB was responsible for
  paper writing, analysis of results, code reviews. LJC was responsible for problem conception and framing, analysis of results, paper writing.
  \end{contributions}

>>>>

Beyond functional annotation, deep learning has enabled significant advances in protein structure prediction \cite{alquraishi2019end, senior2020improved, Yang1496, yang2020improved, du2020energy, rao2021msa}, predicting the functional effects of mutations \cite{riesselman2018deep,rives2019biological,rao2019evaluating,frazer2020large}, and protein design \cite{yang2019machine, mazurenko2019machine, biswas2020low, madani2020progen,elnaggar2020prottrans, anishchenko2020novo, bryant2021deep}. A key departure from traditional approaches is that researchers have started to incorporate vast amounts of raw, uncurated sequence data into model training, an approach which also shows promise for functional prediction \citep{Brandes2021.05.24.445464}.

@@

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are alternative methods for splitting \citep{Brandes2021.05.24.445464,zhou2019cafa, gillis}, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. Such a split is not available for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives \citep{warwick2020benchmarking}.

@@

%startleaveout

\begin{acknowledgements}
  We would like to thank Babak Alipanahi, Jamie Smith, Eli Bixby, Drew Bryant, Shanqing Cai, Cory McLean and Abhinay Ramparasad. The static version of this manuscript uses a template made by Ricardo Henriques.

  TS receives funding from the Wellcome Trust through a Sir Henry Wellcome Postdoctoral Fellowship (210918/Z/18/Z). This work was also supported by the Francis Crick Institute which receives its core funding from Cancer Research UK (FC001043), the UK Medical Research Council (FC001043), and the Wellcome Trust (FC001043). This research was funded in whole, or in part, by the Wellcome Trust [FC001043]. For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.

\end{acknowledgements}
  \begin{contributions}
  TS was responsible for conception, data ingestion, code, analysis, visualisation, UI, paper writing. MLB was
  responsible for visualization, analysis, code, CLI, open sourcing, testing, paper writing. DB was responsible for
  paper writing, analysis of results, code reviews. LJC was responsible for problem conception and framing, analysis of results, paper writing.
  \end{contributions}
